Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-04-23 main.Rmd
up to 3.2

 The methods are implemented in the open source R package gravitas. The functions for creating granularities and exploring the associated time series are consistent with a tidy workflow (@Grolemund2018-po), and the probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk].
bibliography: [bibliography.bib]
preamble: >
  \setlength {\marginparwidth }{2cm}
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

@@

\autoref{fig:linear-time} illustrates the linear time granularities.<!--, each of which are non-repeating in nature and arranged from past to future. If we consider $T$ hours, we will have $\lfloor T/24\rfloor$ days, $\lfloor T/(24*7)\rfloor$ weeks and so on.-->

@@

<!-- CONTENT CUT -->
<!-- These definitions are illustrated in \autoref{fig:groups-finer}, where both day $\trianglelefteq$ week, and day $\preceq$ week, hold true. However, weekends are finer than weeks but do not group into weeks. And weekends neither groups into or is finer than public holidays. Consider another example where day $\trianglelefteq$ month. This relationship is not fully described until we associate it with periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions. -->

For example, $day \trianglelefteq week$ and $day \preceq week$ since every granule of $week$ is the union of some set of granules of $day$ and each $day$ is a subset of a $week$. Consider another example where $day \trianglelefteq month$. This relationship however is incomplete without its association to periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping $(day, month)$ is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions.

@@

\noindent For example, in a Gregorian calendar without leap years we could say $day$ groups periodically into $month$ with the period $P = 365$ and the number of granules of $month$ in each period given by $R=12$. In a Gregorian calendar with leap years, $day$ groups quasi-periodically into $month$ with the <!--The value of $P$ and $R$ would still be 365 and 12 respectively, while the $E_i$s--> exceptions of the time domain corresponding to $29^{\text{th}}$ February of any year.

@@

The grouping operation:
: Let $G_1$ be a full-integer labeled granularity, and $m$ a positive integer. The grouping operation Group$_m(G)$ generates a new granularity $G_2$, by partitioning the granules of $G_1$ into $m$-granule groups and making each group a granule of the resulting granularity. More precisely, $G_2 = \text{Group}_m(G_1)$ is the full-integer labeled granularity such that for each integer $i$,
$$G_2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G_1(j).$$
Example: $\text{minute} = \text{Group}_{60}(\text{second})$; $\text{hour} = \text{Group}_{60}(\text{minute})$

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] data structure. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. <!--In a tsibble, each observation (row) is uniquely identified by index and key.--> A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble.<!--Consistent with the assumption of @Bettini2000-vy, all linear granularities can be generated from the bottom granularity by calendar algebraic operations.-->

Cyclic granularities can be constructed by relating two linear granularities. The mappings between these linear granularities could be either _regular_ or _irregular_ with every granule of coarser linear granularity being made up of same (regular) or different (irregular) number of granules of finer linear granularity. <!--For example, a regular mapping exists between  (where 60 minutes always add up to 1 hour), while an irregular mapping exists between  (as the number of days per month ranges from 28 to 31). A grouping operation in Calendar algebra could lead to a regular mapping, whereas a altering-tick operation could result in an irregular mapping.--> Cyclic time granularities are referred to as **circular** if mappings are regular and **quasi-circular** if the mappings are irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month.

@@

@@

\autoref{fig:circular-dow} depicts the linear (a) and the corresponding stack (b) display of time. Circular granularities day-of-week $C_{B , G}$ and day-of-fortnight $C_{B, H}$ are constructed from linear granularities day ($B$), week ($G$) and fortnight ($H$). The granules that overlap share elements from the underlying time domain in the linear display. The circular granularities $C_{B , G}$ and $C_{B , H}$  <!--= C_{\text{day}, \text{week}}$ representing day-of-week and day-of-fortnight respectively --> consist of repeated pattern of categories $\{0, 1, 2, \dots, 6\}$ and $\{0, 1, 2, \dots, 13\}$ with $P=7$ and $P=14$ respectively. The granules that overlap represents the same category in the stack display. Each circular granularity can use descriptive label mappings, or be simply labelled using the index. <!--Similar to linear granularities, the indexing for each granule of the circular granularity can also be associated with textual representation, called the label.--> Suppose ${L}$ is a label mapping that defines an unique label for each index $l \in \{ 0,1,\dots, (P-1)\}$, then the label mapping $L$ for $C_{B, G}$ can be defined as
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sun}, \text{Mon}, \dots, \text{Sat}\}
$$
or
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sunday}, \text{Monday}, \dots, \text{Saturday}\}
$$
for example.

@@

```{r circular-dow,echo=FALSE,out.width = "100%",fig.cap=" Circular granularities illustrated through linear (a) and stack (b) display of time. The granules that overlap represent same time domain in the linear display and same category in the stack display. Circular granularities day-of-week ($C_{B, G}$) and day-of-fortnight ($C_{B, H}$) have repeating patterns unlike $B$, $G$ or $H$ which gets manifested more clearly through a stack display of time."}
knitr::include_graphics("Figs/circularex.png")
```

@@

### Muliple order-up circular granularities {#sec:all-circular-multiple}

>>>>

 The methods are implemented in the open source R package gravitas. The functions for creating granularities and exploring the associated time series are consistent with a tidy workflow (@Grolemund2018-po), and the probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk].
bibliography: [bibliography.bib]
preamble: >
  \setlength {\marginparwidth }{2cm}
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes,amsthm}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

@@

\autoref{fig:linear-time} illustrates the linear time granularities. \todo[inline]{Expand  into  a paragraph - how does  it illustrate the linear granularities? How does  hour, day,... relate to G...} <!--, each of which are non-repeating in nature and arranged from past to future. If we consider $T$ hours, we will have $\lfloor T/24\rfloor$ days, $\lfloor T/(24*7)\rfloor$ weeks and so on.-->

@@

<!-- CONTENT CUT -->
<!-- These definitions are illustrated in \autoref{fig:groups-finer}, where both day $\trianglelefteq$ week, and day $\preceq$ week, hold true. However, weekends are finer than weeks but do not group into weeks. And weekends neither groups into or is finer than public holidays. Consider another example where day $\trianglelefteq$ month. This relationship is not fully described until we associate it with periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions. -->
\newtheorem*{example}{Example}

\begin{example}
{\rm Let $day \trianglelefteq week$ and $day \preceq week$ since every granule of $week$ is the union of some set of granules of $day$ and each $day$ is a subset of a $week$. Consider another example where $day \trianglelefteq month$. This relationship however is incomplete without its association to periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping $(day, month)$ is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions.}
\end{example}

@@

\begin{example}
{\rm In a Gregorian calendar without leap years we could say $day$ groups periodically into $month$ with the period $P = 365$ and the number of granules of $month$ in each period given by $R=12$. In a Gregorian calendar with leap years, $day$ groups quasi-periodically into $month$ with the <!--The value of $P$ and $R$ would still be 365 and 12 respectively, while the $E_i$s--> exceptions of the time domain corresponding to $29^{\text{th}}$ February of any year.}
\end{example}

@@

The grouping operation:
: Let $G_1$ be a full-integer labeled granularity, and $m$ a positive integer. The grouping operation Group$_m(G)$ generates a new granularity $G_2$, by partitioning the granules of $G_1$ into $m$-granule groups and making each group a granule of the resulting granularity. More precisely, $G_2 = \text{Group}_m(G_1)$ is the full-integer labeled granularity such that for each integer $i$,
$$G_2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G_1(j).$$

\begin{example} 
{\rm minute = G$_{60}(\text{second})$, hour = G$_{60}(\text{minute})$, day = G$_{24}(\text{hour})$.}\todo[inline]{Can you use the same units shown in Fig 1?}
\end{example}

@@

Cyclic granularities represeent cyclical components in time.  Weeks is an example of a cyclical granularity because every 7 days there's another Monday. Cyclic granularities can be constructed by conducting modular arithmetic on two linear granularities. Cycles can be either _regular_, called **circular**, or _irregular_, **quasi-circular**. <!-- with every granule of coarser linear granularity being made up of same (regular) or different (irregular) number of granules of finer linear granularity. For example, a regular mapping exists between  (where 60 minutes always add up to 1 hour), while an irregular mapping exists between  (as the number of days per month ranges from 28 to 31). A grouping operation in Calendar algebra could lead to a regular mapping, whereas a altering-tick operation could result in an irregular mapping. Cyclic time granularities are referred to as **circular** if mappings are regular and **quasi-circular** if the mappings are irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month.-->

@@

Cyclic time granularities can be implemented through the tsibble [@wang2019tsibble] data structure. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. <!--In a tsibble, each observation (row) is uniquely identified by index and key.--> A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble.<!--Consistent with the assumption of @Bettini2000-vy, all linear granularities can be generated from the bottom granularity by calendar algebraic operations.-->\todo[inline]{Would it be worth to start this discussion about  data structures earlier, with the linear granularity, and follow it through when appropriate; or put all discussion on  implementation in the data structures section.}

@@

\autoref{fig:circular-dow} \todo[inline]{C(B,G) should be above C(B,H) in figure, and yuo could add the integers to the stacked sets.} illustrates the linear and the corresponding cyclical granularities. Cyclical granularities can  be considered to be cutting  the  linear  granularity into pieces, and stacking them to match the cycles (as shown in b). $B, G, H$ (day, week, fortnight, respectively) are linear granularities. The circular granularity $C_{B , G}$ (day-of-week)is constructed from $B$ and  $G$. The circular granularity $C_{B, H}$ (day-of-fortnight) is constructed from $B$ and $H$. These overlapping cyclical granularities share elements from the linear granularity. Each of $C_{B , G}$ and $C_{B , H}$ consist of repeated patterns  $\{0, 1, 2, \dots, 6\}$ and $\{0, 1, 2, \dots, 13\}$ with $P=7$ and $P=14$ respectively.

Each circular granularity can use descriptive label mappings. <!--Similar to linear granularities, the indexing for each granule of the circular granularity can also be associated with textual representation, called the label.--> Suppose ${L}$ is a label mapping that defines an unique label for each index $l \in \{ 0,1,\dots, (P-1)\}$, then the label mapping $L$ for $C_{B, G}$ can be defined as
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sun}, \text{Mon}, \dots, \text{Sat}\}
$$
or
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sunday}, \text{Monday}, \dots, \text{Saturday}\}
$$
for example.

@@

```{r circular-dow,echo=FALSE,out.width = "100%",fig.cap="Illustration of circular relative to linear granularities (a). Circular granularities can be considered to be cutting the linear granularity into pieces and stacking them (b). The circular  granularity creates repeated integer sequences."}
knitr::include_graphics("Figs/circularex.png")
```

@@

### Multiple order-up circular granularities {#sec:all-circular-multiple}
